---
title: "PM 560 Lab 3"
author: "Sylwia Lipior"
format: html
embed-resources: true
editor: visual
---

## Lab 3

1.  Read in the data

```{r}

download.file(

    "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",

    destfile = "met_all.gz",

    method   = "libcurl",

    timeout  = 60 )

met <- data.table::fread("met_all.gz")


dim(met)
head(met)
tail(met)
str(met)




```

2.  Check the dimensions, headers, footers. How many columns, rows are there? 2377343 rows, 30 columns

3.  Take a look at the variables

    ```{r}

    str(met)
    ```

4.  Take a closer look at the key variables

    ```{r}
    table(met$year)
    table(met$day)
    table(met$hour)
    summary(met$temp)
    summary(met$elev)
    summary(met$wind.sp)

    met[met$elev==9999.0] <- NA
    summary(met$elev)

    met <- met[temp>-40]
    met2 <- met[order(temp)]
    head(met2)
    ```

5.  Check the data against an external data source.

Remove all the temperatures colder than -15 degrees

```{r}

met <- met[temp>-15]
met2 <- met[order(temp)]
head(met2)

```

6.  Calculate summary statistics

```{r}
elev <- met[elev==max(elev), ]
summary(elev)

cor(elev$temp, elev$wind.sp, use="complete")
cor(elev$temp, elev$hour, use="complete")
cor(elev$wind.sp, elev$day, use="complete")
cor(elev$wind.sp, elev$hour, use="complete")
cor(elev$temp, elev$day, use="complete")
```

```{r setup, include=FALSE}

library(leaflet)
leaflet()


```

7.  Exploratory graphs

Temperature seems to fluctuate a range of around 12 degrees even within a short amount of time.

Periods of higher wind speed seem to come last for at least a few days.

```{r}

hist(met$elev)
hist(met$elev, breaks=100)
hist(met$temp)
hist(met$wind.sp)

library(leaflet)
library(leaflet.extras)

leaflet(elev) %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircles(lat = ~lat, lng = ~lon, opacity = 1, fillOpacity = 1, radius = 100)

  



```

```{r}

library(lubridate)
elev$date <- with(elev, ymd_h(paste(year, month, day, hour, sep= ' ')))
summary(elev$date)

elev <- elev[order(date)]
head(elev)

plot(elev$date, elev$temp, type='l')
plot(elev$date, elev$wind.sp, type='l')
```

8.  Ask Questions

I want to see if there is a relationship between wind speed and temperature at the weather station at the highest elevation.

Assessing the correlation and the scatterplot, there doesn't seem to be a relationship between wind speed and temperature.

```{r}



cor(elev$temp, elev$wind.sp, use="complete")

plot(elev$wind.sp, elev$temp)



```
